Practical Optional Types for Clojure

### 2015-02-19 formal-model.tex
more figures

\begin{figure*}
  \footnotesize
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                &\mbox{Expressions} \\
  \v{} &::=&          \true{} \alt \false{} \alt \trdiff{\nil{}}
                      \alt \trdiff{\class{}}
                      \alt \k{}
                      \alt {\const{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
  {\const{}}           &::=& \classconst \alt \throwconst

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\end{figure*}

@@

@@

@@

\begin{verbatim}
TODO
- example of normal MM usage
- explain descrepency btwn TC syntax and formal
  TR calculs syntax. eg. :object/:filter 
    vs over arrow
\end{verbatim}

@@

\begin{figure}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
                &\mbox{Expressions} \\
  \v{} &::=& \ldots \alt {\emptymap{}}
                &\mbox{Values} \\
  \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
                &\mbox{Types} \\
  \auxhmapsyntax{}\\
\end{altgrammar}
$$
\caption{HMap Terms and Types}
\end{figure}

@@

@@

    \mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Type Hinting and Non-Reflective Terms}
  \label{figure:nonreflectivesyntax}
\end{figure}

@@

\begin{verbatim}
TODO 
- write conversion rules for 
  Java type -> Clojure type
\end{verbatim}

\input{constant}

>>>>

\begin{figure*}
  \footnotesize
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                      \alt {\errorvalv{}}
                &\mbox{Expressions} \\
  \v{} &::=&          \true{} \alt \false{} \alt \trdiff{\nil{}}
                      \alt \trdiff{\class{}}
                      \alt \k{}
                      \alt {\const{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
  {\const{}}           &::=& \classconst \alt \throwconst

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\end{figure*}

@@

    {\TKw}

@@

    {\TError}

@@

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\emptymap{}}
    &\mbox{Values} \\
    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
    &\mbox{Types} \\
    \auxhmapsyntax{}\\
  \end{altgrammar}
  $$
  \caption{HMap Terms and Types}
  \label{figure:hmapsyntax}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\TGetHMap}

    {\TGetAbsent}

    {\TGetHMapPartialDefault}

    {\TAssoc}
  \end{mathpar}
  \caption{HMap Typing Rules}
  \label{figure:hmaptyping}
\end{figure}

@@

It is common in Clojure to use plain immutable hash maps to represent ad-hoc
objects with a known set of keyword entries. We use \HMapliteral{}
types to track known present, known absent and unknown keyword entries
(figure~\ref{figure:hmapsyntax}).

The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
entries and we write \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}} as a fully specified
HMap can assume all entries not in {\mandatory{}} as absent.

\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
To simplify presentation, we only allow the empty map literal and
restrict lookup and extension to keyword keys.

@@

    \mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Non-Reflective Java Interoperability Syntax}
  \label{figure:nonreflectivesyntax}
\end{figure}

@@

\input{constant}
\input{update}
\input{convert-types-figure}
